Water desalination using polyelectrolyte hydrogel: Gibbs ensemble modeling

### 2022-09-15 gibbs.tex

\subsection{Model}

Our model of the gel is a network of 16 linear polymer chains, each consisting of 30 monomer 
units. 
These polymer chains are connected to a diamond-like network by 8 crosslinking units.
This means, there are $\Ngel=16\cdot30 + 8=488$ gel monomers in the simulation box 
(see \reffig{fig:diamond}).
Each monomer unit of the network carries a negative elementary electric charge, thus 
simulating fully ionized polyacidic gel.
Except the particles in the network, the monovalent co- and counter-ions ions, 
$\cl$ and $\na$, are present in simulation box. 
The total electric charge of all the particles in the box is zero, therefore the amount 
of $\na$ ions is bigger than that of $\cl$ by the number of hydrogel units.
The entire simulation box setup employs periodic boundaries.

>>>>

\subsection{Open and closed sytems}

We propose a cascade desalination process to lower supernatant salinity.
%it is worth to note that we are not limited with this process only.
For this process the gel must be compressed/decompressed in an \emph{open} and \emph{closed} system.
By the \emph{open system} we mean that the gel is in thermodynamic equilibrium with huge, effectively \textbf{infinite}, reservoir of an aqueous solution. 
%The gel freely exchanges ions with the reservoir, while their concentration in reservoir (and chemical potential) remains constant.
%, such that there are the same ion's chemical potential, the same temperature, and desired pressure difference between the gel and the reservoir. The number of  $\na$ and $\cl$ ions and gel volume $\Vgel$ are allowed to change.
By the \emph{closed system} we mean that the gel is in equilibrium with a \textbf{finite} reservoir of an aqueous solution. 
In this case the exchange of ions between the gel and the reservoir may affect their concentration in reservoir.%, whereas the  
%again have the same ions chemical potential, the same temperature, and desired pressure difference between the gel and the reservoir. 
%Although, the number of ions and volume of gel and reservoir may vary, 
%total number of $\na$ and $\cl$ ions in the gel and in the reservoir, \ie in the volume $\Vbox$ remains constant.

Due to huge size of the reservoir, the compression of the gel in the \emph{open system} does not affect the reservoir salinity, $\cs =$ $\cna=$ $\ccl=$ $\text{Const}$ (\reffig{fig: open}), whereas the number of ions entrapped in the volume $\Vbox$ changes (\reffig{fig: closed}).
On the contrary, the compression of the gel in the \emph{closed system} changes the salinity in the reservoir, but the total number of $\na$ and $\cl$ ions in the gel and in the reservoir, \ie in the volume $\Vbox$ (\reffig{fig: closed}), remains constant.

%In both cases there are two coexisting systems with constant $\mu P T$ allowed to exchange particles and change volume, which is represented as Gibbs ensemble.

%Unfortunately, there is no universal computationally effective technique to sample pressure difference, volume, and number of particles at the same. 
The mechanical movement and the exchange of ions occur simulataneously in real systems, however, we simulate them alternating each other in stop-run mode in a random order. 
To sample mechanical properties of the gel and reservoir we use MD simulation, whereas to sample the ions distribution between the gel and a reservoir we use MC simulation.
The details of this hybryd computational technique can be found in our previous studies of polyelectrolytes in open system \cite{Rud2020, Rud2022, Landsgesel2020a}.

%In this sense employed techniques is called MCMD.

%To simulate processes in \emph{open system} and  \emph{closed system} two complementary models MC and MD were implemented.

\subsection{Langevin Dynamics}

We model the gel as a network of 16 linear polymer chains, each consisting of 30 monomer units. 
These polymer chains are connected to a diamond-like network by 8 crosslinking units.
This means, there are $\Ngel=16\cdot30 + 8=488$ gel monomers in the simulation box 
(see \reffig{fig:diamond}).
Each monomer unit of the network carries a negative elementary electric charge, thus 
simulating fully ionized polyacidic gel.
Except the particles in the network, the monovalent co- and counter-ions ions, 
$\cl$ and $\na$, are present in simulation box. 
The total electric charge of all the particles in the box is zero, therefore the amount 
of $\na$ ions is bigger than that of $\cl$ by the number of hydrogel units, $\Ngel$.
The solvent assumed to be water, which is modeled implicitly via setting up dielectric permittivity $\epsilon=80$.

We use three types of interactions, \ie bonded interactions, short-range non-bonded interactions and electrostatic interactions.

We use Finite extension nonlinear elastic (FENE) potential~\refeq{eq:fene} to account for the bonding interactions.
\begin{equation}\label{eq:fene}
    V_\mathrm{FENE}(r) = -\frac{1}{2} K \Delta r_\mathrm{max}^2\ln \left[ 1 - \left(\frac{r-r_0}{\Delta r_\mathrm{max}} \right)^2 \right ],
\end{equation}
where $K$ is the magnitude of symmetric interaction between two segments,
$\Delta r_{\mathrm{max}}$ is the maximal stretching length of the bond and $r_0$ is the equilibrium bond length.
In our simulations we set $K = 10 \kT/\sigma^{2}$, $\Delta r_{\mathrm{max}} = 2 \sigma$ and $r_0 = 1.0\sigma$~\cite{Jin2007}.

The non-bonded interactions are modelled by the truncated Lennard-Jones interaction potential.
This potential imposes strong repulsion between all particles at short distances:
\begin{equation}
    \begin{split}\label{eq:LJ}
        V_\mathrm{LJ}(r) =
        \begin{cases}
            4 \varepsilon \left( \left(\cfrac{\sigma}{r}\right)^{12}
            - \left(\cfrac{\sigma}{r}\right)^6\right)
            & \mathrm{if~} r < r_\mathrm{cut}\\
            0
            & \mathrm{elsewhere}
        \end{cases},
    \end{split}
\end{equation}
where $r$ is the interparticle distance,
$\sigma = 0.35 \mathrm{nm}$ is a chosen characteristic size of the particles,
$\varepsilon$ is the depth of the potential well in $\kT$ units and $r_\mathrm{cut}$ is the cut-off distance beyond which the potential is set zero.

We model the long range electrostatic interactions by the Coulomb potential:
\begin{equation}
    V_{\mathrm{EL}}=\lb\kT \cfrac{q_{1}q_{2}}{r},
\end{equation}
where $\lb$ is Bjerrum length; $\lb = 2\sigma = 0.7 \mathrm{nm}$ which corresponds to water the Bjerrum length in water at $T=300 \mathrm{K}$.

We use Langevin thermostat, \ie we add two additional terms to force in equation of motion
\begin{equation}
f_i =  -\gamma v_i(t)+\sqrt{2\gamma \kT }\eta_i(t),
\end{equation}
where the first term correspond to constant friction with $\gamma$ being a friction coefficient, and the second one corresponds to random thermal force with $\eta_i$ being a normally distributed random number (see for details \cite{Grest1986}).

Integrating the Langevin motion equation for all the particles we sample total energy of the system and pressure \cite{essmann1995}.

\subsubsection{Monte Carlo sampling in open system}

Monte Carlo scheme for sampling the exchange of ions in \emph{open system} is based on the formula for free energy of grand canonical ensemble.

\begin{equation}
    \Omega=E-TS+\sum_i\mu_i N_i\label{eq:Omega-GC}
\end{equation}
where $E$ is internal energy, $T$~--- temperature, $S$~--- entropy, $N_i$ is the number of ions of type $i\in\{\na,\ \cl\}$.

%Our model system represents a finite volume $V$ which exchanges particles with the bath of infinite volume.
%The infinity of the volume means, that the particle exchange between the system and the bath does not influence on the chemical potentials in the bath.
%Such ensemble is called grand canonical ensemble; the free energy in this ensemble, the Landau potential, is
The entropy $S$ is given by the Boltzmann formula~\cite{Nagle2004}:
\begin{equation}
    S=\kB\sum_i\ln\frac{V^{N_i}}{N_i!}\label{eq:entropy}
\end{equation}
%\todo{for partiles without interactions}
It accounts for two contributions:
\begin{enumerate}
    \item the combinatorial entropy $S_{c}=-\sum_i\kB\ln N_i!$ which reflects the freedom of choice among $N$ particles
    \item the mixing entropy $S_{m}=\kB \sum_i N_i\ln V$ which reflects the freedom to place the chosen particle randomly within the simulation box. 
\end{enumerate}
$V$ is the unitless volume, \ie the volume measured in units of $\sigma^3$. 
Thus \refeq{eq:Omega-GC} turns to 
%\todo{this is not reason, moreover A.10 is already wrong as for units}
\begin{equation}
    \Omega=E-\kT\ln\prod_i\frac{V^N_i}{N_i!}+\sum_i\mu_i N_i
\label{eq: Omega_GC}
\end{equation}
The change of free energy associated with an exchange of ion pair is
\begin{equation}
    \Delta\Omega=\kT\ln\prod_i\left(V^{\xi}\frac{N_i!}{\left(N_i+\xi\right)!}\right)+\xi\mu_i+\Delta E
\label{eq: DeltaG GC}
\end{equation}
where $\xi=+1$ stands for insertion of an ion pair and $\xi=-1$ for deletion.
In general $\xi$ can be any number, but in case when  $\xi = \pm 1$ the previous equation gets simplified 
\begin{equation}
    \Delta\Omega=\kT\ln V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}+\xi\mu_i+\Delta E
\label{eq: DeltaG GC_simpl}
\end{equation}
where $\theta$ is the Heaviside function, $\theta(\xi) = 1, \mathtt{if}\  \xi>0$ , $\theta(\xi) = 0, \mathtt{if}\  \xi<0$

The procedure of the Monte Carlo Sampling is the following \cite{Frenlkel2002_book}
\begin{enumerate}
	\item we propose the new configuration of the system by insertion (or deletion) of an ion pair, $\xi=\pm1$
	\item we accept the new configuration if
	\begin{equation}
        \mathcal{R}^{\xi}<e^{\Delta\Omega/\kT}=V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}e^{\left(\Delta E+\xi\mu\right)/\kT}\label{eq: GC acceptance}
	\end{equation}
\end{enumerate}
where $\mathcal{R}$ is uniformly distributed random number in range between $0$ and $1$.

\subsection{Monte Carlo sampling in closed system}
In the same way we derive the acceptance criterion for sampling the closed system (Gibbs ensemble)

The free energy of the Gibbs ensemble is a sum of free energies of the gel and the external volume.

\begin{equation}
    \Omega=E\gel-TS\gel + E\out-TS\out \label{eq:Omega-GC}
\end{equation}

The entropy $S$ is given by the Boltzmann formula~\cite{Nagle2004}:
\begin{equation}
    S=\kB\sum_i\ln\frac{V\gel^{N\gel_i}}{N_i!}\label{eq:entropy}
\end{equation}
%\todo{for partiles without interactions}
It accounts for two contributions:
\begin{enumerate}
    \item the combinatorial entropy $S_{c}=-\sum_i\kB\ln N_i!$ which reflects the freedom of choice among $N$ particles
    \item the mixing entropy $S_{m}=\kB \sum_i N_i\ln V$ which reflects the freedom to place the chosen particle randomly within the simulation box. 
\end{enumerate}
$V$ is the unitless volume, \ie the volume measured in units of $\sigma^3$. 
Thus \refeq{eq:Omega-GC} turns to 
%\todo{this is not reason, moreover A.10 is already wrong as for units}
\begin{equation}
    \Omega=E-\kT\ln\prod_i\frac{V^N_i}{N_i!}+\sum_i\mu_i N_i
\label{eq: Omega_GC}
\end{equation}
The change of free energy associated with an exchange of ion pair is
\begin{equation}
    \Delta\Omega=\kT\ln\prod_i\left(V^{\xi}\frac{N_i!}{\left(N_i+\xi\right)!}\right)+\xi\mu_i+\Delta E
\label{eq: DeltaG GC}
\end{equation}
where $\xi=+1$ stands for insertion of an ion pair and $\xi=-1$ for deletion.
In general $\xi$ can be any number, but in case when  $\xi = \pm 1$ the previous equation gets simplified 
\begin{equation}
    \Delta\Omega=\kT\ln V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}+\xi\mu_i+\Delta E
\label{eq: DeltaG GC_simpl}
\end{equation}
where $\theta$ is the Heaviside function, $\theta(\xi) = 1, \mathtt{if}\  \xi>0$ , $\theta(\xi) = 0, \mathtt{if}\  \xi<0$

The procedure of the Monte Carlo Sampling is the following \cite{Frenlkel2002_book}
\begin{enumerate}
	\item we propose the new configuration of the system by insertion (or deletion) of an ion pair, $\xi=\pm1$
	\item we accept the new configuration if
	\begin{equation}
        \mathcal{R}^{\xi}<e^{\Delta\Omega/\kT}=V^{2\xi} \prod_i \left(N_i+\theta(\xi)\right)^{-\xi}e^{\left(\Delta E+\xi\mu\right)/\kT}\label{eq: GC acceptance}
	\end{equation}
\end{enumerate}
where $\mathcal{R}$ is uniformly distributed random number in range between $0$ and $1$.

We model the ionization equilibrium by means of sampling the reaction ensemble~\cite{Smith1994, Turner2008a}.
In the reaction ensemble we perform chemical reactions as simultaneous insertion, deletion or identity change of particles according to the reaction equation.
For example we assume the ionization of an acidic unit which proceeds according to the reaction equation
\begin{equation}
    \mathrm{AH}\stackrel{K}{\leftrightarrows}\A+\h\label{eq: acid reaction}
\end{equation}
We perform the forward direction of this reaction
by inserting a $\h$ particle and replacing $\AH$ by $\A$.
We perform the reverse direction of this reaction
by deleting a $\h$ particle and replacing $\A$ by $\AH$.

In order to account for the particles creation and deletion
we need to introduce standard states for individual reacting species.
The creation of a particle then means that the particle is removed from the standard state
and inserted into the studied system, and vice versa for deletion.
Each species in its standard state has the \emph{standard} chemical potential, $\mu^\ominus_i$.
The choice of standard state for each component is arbitrary
and we usually define it as pure component at standard pressure $10^5$ Pa
and concentration $1 \molperl$~\cite{IUPAC}.

Thus we rewrite the formula for free energy \refeq{eq:Omega-GC}
accounting for particles creation and deletion as
\begin{equation}
    \Omega=E-TS+\sum_i\left(\mu_i-\mu^\ominus_i\right) N_i\label{eq:Omega-RE}
\end{equation}
where the last term in \refeq{eq:Omega-RE} stands for the exchange of particles between the studied system and the standard states.

Let's have the reaction extent $\xi = (N_i - N_{i,0})/\nu_i$
where $N_i$, resp. $N_{i,0}$ is the actual, resp. initial number of the $i$-th particle species,
$\nu_i$ is its stoichiometric coefficient.
and $i$ runs over individual species involved
in the chemical reaction \refeq{eq: acid reaction} $i \in (\h,\ \A,\ \AH)$. 
The stoichiometric coefficients are positive for reaction products
and negative for reactants: $\nu_\h=1$, $\nu_\A=1$, $\nu_\AH=-1$.
Therefore we take $\xi = +1$ for the forward reaction direction
and $\xi = -1$ for the reverse reaction direction.
Then we can write the change of the free energy of the studied system during one reaction step as
\begin{equation}
\Delta\Omega_{\mathrm{re}}=\kT\ln\left(\prod_i V^{\nu_i\xi}\frac{N_i!}{\left(N_i+\nu_i\xi\right)!}\right)+
\xi\left(\sum_i\nu_i\mu_i-\sum_i\nu_i\mu_i^\ominus\right)+\Delta E
\label{eq: DeltaG RE one}
\end{equation}
%\begin{multline}
%    \Delta\Omega_{\mathrm{re}}=\kT\ln\left(\prod_i V^{\nu_i\xi}\frac{N_i!}{\left(N_i+\nu_i\xi\right)!}\right)+\\
%    \xi\left(\sum_i\nu_i\mu_i-\sum_i\nu_i\mu_i^\ominus\right)+\Delta E
%\label{eq: DeltaG RE one}
%\end{multline}

We write $\sum_i\nu_i \mu_i = 0$
for the chemical reaction in equilibrium.

The term $\sum_i\nu_i\mu_i^\ominus$ is the standard Gibbs free energy change
of the reaction \refeq{eq: acid reaction}.
which defines the equilibrium reaction constant \cite{Atkins} as
\begin{equation}
    K = e^{-\sum_i\nu_i \mu_i^\ominus / \kT}
\end{equation}

Finally, the \refeq{eq: DeltaG RE one} translates to
\begin{equation}
    \Delta\Omega_{\mathrm{re}}=\kT\ln\left(K^\xi\prod_i V^{\nu_i\xi}\frac{N_i!}{\left(N_i+\nu_i\xi\right)!}\right)+\Delta E
\label{eq: DeltaG RE two}
\end{equation}

\subsection{Ion pairs insertions\label{sec: ion insertions}}
We treat the exchange of mobile ions with the reservoir as simultaneous insertion or deletion
of ion pairs or sets in the simulation box.
We insert/delete either $\na$ and $\cl$ ion pair or the two $\cl$ ions together with one $\ca$ ion.
Thus we model the insertion as:
\begin{eqnarray}
    \na +  \cl &\leftrightarrows& \emptyset\label{eq:nacl} \\
    \ca + 2\cl &\leftrightarrows& \emptyset \label{eq:cacl}
\end{eqnarray}
with the corresponding equilibrium reaction constants, reflecting the free energy change during the insertion
\begin{eqnarray*}
    K_\mathrm{NaCl}     = \exp\left(\left(\mu_\na+ \mu_\cl \right) /\kT\right), \\
    K_{\mathrm{CaCl}_2} = \exp\left(\left(\mu_\ca+2\mu_\cl \right /\kT\right))
\end{eqnarray*}
%$K_\mathrm{NaCl}= \exp\left(\mu_\na+\mu_\cl \right)$, $K_{\mathrm{CaCl}_2} = \exp\left(\mu_\ca+2\mu_\cl \right)$

the Monte Carlo in the \emph{open system} box with the gel and the monovalent co- and counter-ions ions, 
$\cl$ and $\na$ is allowed to exchange particles with infinite reservoir with constant chemical potential.

\dots

In the \emph{closed system} system two boxes one with the gel and the other with ions pairs only iare allowed to exchange particles.
